Fast inhibitory loop explains functional map development with short-range inhibition.

### 2017-01-10 paper.tex
changes

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{insval}
\usepackage{natbib}

@@

\title{Fast inhibitory loop explains functional map development with short-range inhibition.}
\author{J\'an Antol\'ik}

\begin{document}
 \maketitle
\loadkeyvaluestore{/home/jan/Doc/Papers/fast_inh_paper/key_value_store.py}

@@

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly range of brain functions including working memory \cite{Amit1995,Durstewitz2000},
orientation tuning \cite{Somers1995,Ben-Yishai1995}, and functional map development \cite{VonderMalsburg1973,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, and previous studies have shown that a specific arrangement of the effective inhibitory and excitatory 
interactions is required: inhibition has to have longer range than excitation (a so called Mexican-Hat profile) \cite{Muir2014}.
This is, however, in stark contrast to the know anatomical arrangement: the excitatory neurons superficial layers tend to form 
long-range arborizations spanning multiple columns while the axons of majority of inhibitory neurons are confined to area
only several hundreds of micrometers in diameter \cite{Buzas2006,Budd2001}.

As pointed out by Muir et al. in their recent study \cite{Muir2014}, to solve this apparent conflict, previous topologically organized models cortical competition have either relied on the anatomically unsupported Mexican-hat profile of lateral interactions \cite{VonderMalsburg1973,CMVC}, or relied on other biologically unrealistic properties, such as selective targeting of inhibitory neurons by long range excitatory connections \cite{law:phd09,Rutishauser2012} or instantaneous synaptic transmission coupled with omission of recurrent inhibition \cite{Kang2003,Levy2011}. Model analysis by Muir et al. \cite{Muir2014} showed that the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multi-synaptic network effects are negligible. In conclusion, currently no satisfactory explanation of how topological functional organization dev?lops in cortical networks that is consistent with the present anatomical findings exists.

All previous topologically organized models of cortical competition have either assumed instantaneous synaptic transmission \cite{VonderMalsburg1973,Kang2003,Levy2011,Grabska-Barwinska2008}} or uniform transmission delays across the different intra-cortical connections \cite{CMVC,Antolik2011}. This is also the case of the analysis performed by Muir et al. \cite{Muir2014} which questions the ability of the accepted cortical architecture to explain cortical competition. However, a recent study by Ohana et al. \cite{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelArchitecture/drawing.png}
\caption{General model architecture. (A) Inhibitory cortical population corresponds to a regular lattice of units in cortical space. (B) As in A but 
for excitatory population. (C) Thalamic neurons are modeled as simple difference-of-Gaussian filter followed by a threshold-linear trasfer function. 
(D) The intra-cortical connectivity between the excitatory and inhibitory populations and their transmission delays. The intra-cortical connectivity is the only aspect of the general architecture presented here that changes between the diffrent model variants explored in this study and each variant is detail in figure \ref{fig:model_variants}.
(F) The centers of RFs of the LGN neurons form a regular latice accross the visual space covered by the model. This retinotopic mapping of connection fields between layers of the model is also maintained in the thalamo-cortical projection (lines between D and F)}
\label{fig:model_architecture}
\end{figure}

@@

In sections \ref{sec:SM1} through \ref{sec:SM3} we use GCAL model \cite{Stevens2013}  which is the most advanced variant of LISSOM 
(Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen et al. \cite{CMVC}, 
which itself is based on earlier Self-Organizing Map models \cite{Kohonen1982}.

For clarity and consistency the following model description closely follows the methodology sections in our 
previous work \cite{Stevens2013,Antolik2011}. The architecture of the 3 GCAL variants presented in this 
paper is depicted in figure \ref{fig:fig:model_architecture} and figure \ref{model_variants}A-C. The models are implemented in the Topographica simulator, freely available at www.topographica.org. 
Each of the GCAL models consists of set of sheets of model neural units, corresponding to: (1) the photoreceptors, (2) the combined effects of the retinal ganglion cells (RGC) and ON and OFF LGN cells. Furthermore each model has one V1 sheet of units with direct excitatory and inhibitory interactions (section \ref{sec:SM1}), or two 
V1 sheets, one corresponding to excitatory and one to inhibitory neurons (sections \ref{sec:SM2} and \ref{sec:SM3}).
The model sheet is a 2D array of computational elements (called units or, loosely, neurons), with activation and plasticity equations as 
described below and referenced by a coordinate system we will refer to as sheet coordinates, where the center of the sheet is designated (0.0,0.0).  
The number of units simulated in each sheet is determined by the density of units per unit length in both sheet dimensions. 
All cortical sheets have nominal dimensions \insval{cortex_size}$\times$\insval{cortex_size} in sheet coordinates. The sizes of the
RGC/LGN (\insval{lgn_size}$\times$\insval{lgn_size}) and photoreceptor (\insval{visual_field_size}$\times$\insval{visual_field_size}) sheets
were chosen to ensure that each unit in the receiving sheet has a complete set of connections, thus minimizing edge effects in the
RGC/LGN and V1 sheets. The density of units per $1.0\times$1.0 area is \insval{cortex_density}$\times$\insval{cortex_density} 
for the photoreceptor and RGC/LGN ON and OFF sheets, and \insval{cortex_density}$\times$\insval{cortex_density} for both cortical sheets.

@@

The ON/OFF units are called RGC/LGN units because they represent the entire pathway between the retinal photoreceptors
and V1, including the retinal ganglion cells, LGN cells, and the connection pathways. The activation level for a unit at position $j$
in an RGC/LGN sheet at time $t$ is defined as:
%%
\begin{equation}
\eta_{j}(t)=f\left(\frac{A_{j}(t)}{c + \gamma_{L}\sum_{k}A_{k}(t)l_{kj}}\right)
\end{equation}
%%
where
%%
\begin{equation}
A_{j}(t) = \gamma_{F}\sum_{i\in F_j}\Psi_{i}(t)\omega_{ij}
\end{equation}
%%
The activation function $f$ is a half-wave rectifying function that ensures positive activation values, constant $\gamma_{F}=\insval{retina_to_lgn_strength}$ defines the overall strength of the afferent connections from the retina, constant $\gamma_{L}=\insval{lateral_lgn_strength}$ defines the strength of the lateral connections within the RGC/LGN sheet, constant $c=\insval{lateral_lgn_gain}$ defines the slope of the gain, $\Psi_{i}$ is the activation of unit $i$ taken from the set of photoreceptors
from which RGC/LGN unit $j$ receives input (its connection field $F_j$), $\omega_{ij}$ is the connection weight from unit $i$ in the
retina to unit $j$ in the RGC/LGN, and $l_{kj}$ is the lateral connection weight from RGC/LGN unit $k$ to RGC/LGN unit $j$.
Weights from the photoreceptors to units in the ON and OFF channels are set to fixed strengths with a difference-of-Gaussians
kernel ($\sigma_{center}=\insval{lgn_center_sigma}$, $\sigma_{surround}=\insval{lgn_surround_sigma}$, in sheet dimensions), with ON connection fields having a positive center and a negative surround and vice versa for OFF.  The lateral RGC/LGN weights are 2D Gaussians with kernel size $\sigma=\insval{lgn_lateral_sigma}$. The center of the afferent and lateral connection field of each ON/OFF unit is mapped to the location in the photoreceptor and lgn sheet corresponding to the location of that unit in sheet coordinates, making all these projections retinotopic.

@@

Units in the cortical sheets each receive three types of projections represented as matrices of weights: afferent excitatory $(p=A)$, 
lateral excitatory $(p=E)$ and lateral inhibitory $(p=I)$.  The contribution $C_{jp}$ to the activation of unit $j$ in a cortical sheet from each projection $p$ at time $t$ is given by: 
\begin{equation} 
C_{jp}(t+\delta t)=\sum_{i\in F_{jp}}\Psi_{i}(t)\omega_{pij} 
\end{equation}

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. Table \ref{table:param} shows the strength, initial Gaussian kernel spatial extent, and the cut-off distance values for all projections for the first form of the GCAL model examined in section \ref{sec:SM1}. Any modifications to these base parameters 
in the GCAL model variants examined in sections \ref{sec:SM2} and \ref{sec:SM3} are then reported in the respective sections.

@@

where $\xi=\insval{homeostatic_learning_rate}$ is the time constant of the threshold adaptation,
$\mu=\insval{target_activity}$ is a constant defining the target average activity, and $\tilde{\Psi}$ is 
the recent average activity of the unit:

@@

@@

where $\beta_{p}$ is the Hebbian learning rate for the connection fields in the two afferent projections from RGC/LGN $p \in \{\mathrm{ON},\mathrm{OFF}\}$. I.e., the afferent weights from RGC/LGN are normalized jointly.  Learning rate parameters are specified as
a fixed value $\iota_{p}=\insval{learning_rate}$ for each projection, and then the unit-specific values used in the equation above are calculated as
$\beta_{p}=\frac{\iota_{p}}{\upsilon_{p}}$, where $\upsilon_{p}$ is the number of connections per connection field in projection $p$.

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelVariants/drawing.png}
\caption{The four considered intra-cortical connectivty variants. (A-C) Are modelled using firing rate models of neural units with instantenous 
translation of inputs into membrane-potential (see section \ref{sec:gcal}). 
(A) Variant 1, assuming only local equaly wide excitatory and inhibitory connectivity, and ignoring direct inhibitory to inhibitory interactions. (B) Variant 2, as in A but with added direct inhibitory to inhibitory connections of equal extent as the other excitatory and inhibitory projections. (C) Variant 3, as in A but with added long-range excitatory connections represented based on Buz\'as \etal\,\cite{Buzas2006} as a second wider gaussian. (D) The same as A but modelled with neural units that take into account membrane time-cosntant (see section \ref{sec:rate}).
The transmission delay of the inhibitory to excitatory projections was varied in the experiments presneted in section \ref{sec:SM4}.}
\label{fig:model_variants}
\end{figure}

\subsection{Rate model} \label{sec:rate}

The architecture of this model is identical to the GCAL models presented above (see figure \ref{fig:fig:model_architecture}) 
with the exception of the equation ? describing the neural model which is replaced with equation taking into account the 
membrane time constant:

\begin{equation}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}

As a consequence the evolution of the network dynamics have to be simulated at higher resolution. We chose the update step to be ?ms, and we 
let the activity in the model to settle for ?ms, thus resulting in ? settling steps as opposed to the \ref{} of the GCAL model.

figure \ref{fig:fig:model_architecture} and figure \ref{model_variants}A-C

@@

Model orientation maps are calculated based on the vector average method \cite{CMVC}. We first determine the preferred frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean preferred spatial frequency across all cortical neurons as the
value for the spatial frequency parameter across all subsequent analysis. Next, sine grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180$\circ$), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

In this article we will proceed through multiple gradually more complex models of orientation development, addressing several 
of the major issues with modeling of this phenomena, eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range inhibition can support development of cortical functional organization. We will use two different computational abstractions to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM family of models \cite{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain further in section \ref{sec:SM4}, some simplifications made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.

>>>>

\documentclass[utf8]{frontiersSCNS}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{insval}

\linenumbers

@@

%\title[Running Title]{Fast inhibitory loop explains functional map development with short-range inhibition.}
%\author{J\'an Antol\'ik}

%\begin{document}
%\maketitle

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Antol\'ik} 
\def\Authors{J\'an Antol\'ik\,$^{1,*}$}
\def\Address{$^{1}$ Unit\'e de Neurosciences, Information et Complexit\'e,CNRS UPR 3293, Gif-sur-Yvette, France }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{J\'an Antol\'ik}
\def\corrEmail{antolikjan@gmail.com}

\begin{document}

\onecolumn
\firstpage{1}

\title[Running Title]{Fast inhibitory loop explains functional map development with short-range inhibition.}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated
\extraAuth{}

\maketitle

\begin{abstract}

\section{}

Competitive interactions are believed to underlie variety of cortical processing, ranging from memory formation, attention and 
development of cortical functional organization (e.g. development of orientaion maps in primary visual cortex). 
In the latter case, the competitive interactions happen along the cortical 
surface, with local populations of neurons reinforcing each other while competing with those displaced more distally. This 
specific configuration of lateral interactions is however in stark contrast with the known properties of the antatomical substrate, i.e.
excitatory connections (mediating reinforcment) having longer reach than inhibitory ones (mediating competition). No 
satisfactory biologically plausible resolution of this conflict between anatomical measures, and assumed cortical function has been proposed.
Recently a specific pattern of delays between different types of neurons in cat cortex has been discovered, where direct mono-synaptic
excitation has approximately the same delay, as the combined delays of the disynaptic inhibitory interactions between excitatory neurons 
(i.e. the sum of delays from excitatory to inhibitory and from inhibitory to excitatory neuron). Here we show that this spefic pattern
of delays represents a biologically plausible explanation for how short-range inhibition can support competitive interactions in cortex
that underlie development of orientation maps in primary visual cortex. We demonstrate this statement analytically under simplyfing conditions, 
and subsequently show using network simulations that development of orientation maps is preserved when long-range excitation, 
direct inhibitory to inhibitory interactions, and moderate inequality in the delays between excitatory and inhibitory pathways is 
added.

\tiny
\keyFont{ \section{Keywords:} primary visual cortex, rate model, cortical functional development, hebbian learning, cortical horizontal connectivity, orientation map} 
\end{abstract}

\loadkeyvaluestore{/Users/jan/Documents/doc/Papers/fast_inh_paper/key_value_store.py}

@@

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly range of brain functions including working memory \citep{Amit1995,Durstewitz2000},
orientation tuning \citep{Somers1995,Ben-Yishai1995}, and functional map development \citep{VonderMalsburg1973,CMVC,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, whereby local populations mutually rainforce each other via excitatory connections (short-range 
excitation) while long-range inhibition facilitates competition between the local populations and stabilizes the activity in the 
netowrk. Such so called Mexican-hat arrangement of recurrent interactions, however, is in stark contrast to the know anatomical arrangement
of cortical cicuitry: the excitatory neurons (especially in superficial layers) tend to form long-range arborizations spanning 
multiple columns while the axons of majority of inhibitory neurons are confined to area only several hundreds of micrometers 
in diameter \citep{Buzas2006,Budd2001}.

As pointed out by Muir \etal\, in their recent study \citep{Muir2014}, to solve this apparent conflict, previous topologically organized models of cortical competition have either relied on the anatomically unsupported Mexican-hat profile of lateral interactions \citep{VonderMalsburg1973,CMVC}, or relied on other biologically unrealistic properties, such as selective targeting of inhibitory neurons by long range excitatory connections \citep{law:phd09,Rutishauser2012} or instantaneous synaptic transmission coupled with omission of recurrent inhibition \citep{Kang2003,Levy2011,Grabska-Barwinska2008}.

A common hypothesis about how to reconcile Mexican-hat lateral interacions with anatomical reality is that the range of the effective inhibitory influence of an excitatory neuron onto other excitatory neurons (via the disynaptic pathway from exctiatory to inhibitory to excitatory neurons - the disynaptic inhibition) will correspond to the cumulative reach of the direct excitatory and the direct inhibitory connections. The effective
inhibitory interactions between excitatory neurons will thus have longer range than the direct excitatory connections, supporting the use
of Mexican-hat lateral interaction in single population models (i.e. models without explicit sepparation of excitatory and
inhibitory populations). Even though very intuitive, this explanation has never been explicitly demonstrated, and  
it ommits the fact that under the reasonable null-hypothesis of equal transmission delays on all connections, 
the recurrent disynaptic inhibition will lag the direct recurrent excitation. As it turns out this is crucial, 
as recent model analysis by Muir \etal\,\citep{Muir2014} shows that, under the assumption of uniform tranmission delays, the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multi-synaptic network effects are negligible, effectively rejecting the disynaptic explanation behind Maxican-hat interaction. In conclusion, currently no satisfactory explanation of how topological functional organization develops in cortical networks that is consistent with the present anatomical findings exists.

As we will show in this study, the nature of transmission delays between different neural type (excitatory,inhibitory) combinations
holds the key to resolving this long standing open question. Recent study by Ohana \etal\,\citep{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelArchitecture/drawing.png}
\caption{General model architecture. (A) Inhibitory cortical population corresponds to a regular lattice of units in cortical space. (B) As in A but 
for excitatory population. (C) Thalamic neurons are modeled as simple difference-of-Gaussian filter followed by a threshold-linear trasfer function. 
(D) The intra-cortical connectivity between the excitatory and inhibitory populations and their transmission delays. The intra-cortical connectivity is the only aspect of the general architecture presented here that changes between the diffrent model variants explored in this study and each variant is detail in figure \ref{fig:model_variants}.
(F) The centers of RFs of the LGN neurons form a regular latice accross the visual space covered by the model. This retinotopic mapping of connection fields between layers of the model is also maintained in the thalamo-cortical projection (lines between D and F). These thalamo-cortical projections are the only connections in the model that undergo hebbian adaptation during the simulated visually driven development.}
\label{fig:model_architecture}
\end{figure}

@@

In sections \ref{sec:SM1} through \ref{sec:SM3} (model 1, 2 and 3; see figure \ref{fig:model_variants}) we use GCAL model \citep{Stevens2013}  which is the most advanced variant of LISSOM (Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen \etal\,\citep{CMVC}, which itself is based on earlier Self-Organizing Map models \citep{Kohonen1982}.

For clarity and consistency the following model description closely follows the methodology sections in our 
previous work \citep{Stevens2013,Antolik2011}. The architecture of the 3 GCAL variants presented in this 
paper is depicted in figure \ref{fig:model_architecture} and figure \ref{fig:model_variants}A-C. The models are implemented in the Topographica simulator, freely available at www.topographica.org. 
Each of the GCAL models consists of set of sheets of model neural units, corresponding to: (1) the photoreceptors, (2) the combined effects of the retinal ganglion cells (RGC) and ON and OFF LGN cells. Furthermore each model has one V1 sheet of units with direct excitatory and inhibitory interactions (model 1; see figure \ref{fig:model_variants}), or two 
V1 sheets, one corresponding to excitatory and one to inhibitory neurons (model 2 and 3; see figure \ref{fig:model_variants}).
The model sheet is a 2D array of computational elements (called units or, loosely, neurons), with activation and plasticity equations as 
described below and referenced by a coordinate system we will refer to as sheet coordinates, where the center of the sheet is designated (0.0,0.0).  
The number of units simulated in each sheet is determined by the density of units per unit length in both sheet dimensions. 
All cortical sheets have nominal dimensions \insval{cortex_size}$\times$\insval{cortex_size} in sheet coordinates. The sizes of the
RGC/LGN (\insval{lgn_size}$\times$\insval{lgn_size}) and photoreceptor (\insval{visual_field_size}$\times$\insval{visual_field_size}) sheets
were chosen to ensure that each unit in the receiving sheet has a complete set of connections, thus minimizing edge effects in the
RGC/LGN and V1 sheets. The density of units per $1.0\times$1.0 area is \insval{cortex_density}$\times$\insval{cortex_density} 
for the photoreceptor and RGC/LGN ON and OFF sheets, and \insval{cortex_density}$\times$\insval{cortex_density} for both cortical sheets.

@@

The ON/OFF units are called RGC/LGN units because they represent the entire pathway between the retinal photoreceptors
and V1, including the retinal ganglion cells, LGN cells, and the connection pathways. The activation level for a unit at position $j$
in an RGC/LGN sheet at time $t$ is defined as:
%%
\begin{equation}
\eta_{j}(t)=f\left(\frac{A_{j}(t)}{c + \gamma_{L}\sum_{k}A_{k}(t)l_{kj}}\right)
\end{equation}
%%
\noindent where
%%
\begin{equation}
A_{j}(t) = \gamma_{F}\sum_{i\in F_j}\Psi_{i}(t)\omega_{ij}
\end{equation}
%%
The activation function $f$ is a half-wave rectifying function that ensures positive activation values, constant $\gamma_{F}=\insval{retina_to_lgn_strength}$ defines the overall strength of the afferent connections from the retina, constant $\gamma_{L}=\insval{lateral_lgn_strength}$ defines the strength of the lateral connections within the RGC/LGN sheet, constant $c=\insval{lateral_lgn_gain}$ defines the slope of the gain, $\Psi_{i}$ is the activation of unit $i$ taken from the set of photoreceptors
from which RGC/LGN unit $j$ receives input (its connection field $F_j$), $\omega_{ij}$ is the connection weight from unit $i$ in the
retina to unit $j$ in the RGC/LGN, and $l_{kj}$ is the lateral connection weight from RGC/LGN unit $k$ to RGC/LGN unit $j$.
Weights from the photoreceptors to units in the ON and OFF channels are set to fixed strengths with a difference-of-Gaussians
kernel ($\sigma_{center}=\insval{lgn_center_sigma}$, $\sigma_{surround}=\insval{lgn_surround_sigma}$, in sheet dimensions), with ON connection fields having a positive center and a negative surround and vice versa for OFF.  The lateral RGC/LGN weights are 2D Gaussians with kernel size $\sigma=\insval{lgn_lateral_sigma}$. The center of the afferent and lateral connection field of each ON/OFF unit is mapped to the location in the photoreceptor and lgn sheet corresponding to the location of that unit in sheet coordinates, making all these projections retinotopic.

@@

Units in the cortical sheets each receive three types of projections represented as matrices of weights: afferent excitatory $(p=A)$, 
lateral excitatory $(p=E)$ and lateral inhibitory $(p=I)$.  The contribution $C_{jp}$ to the activation of unit $j$ in a cortical sheet from each projection $p$ at time $t$ is given by: 
\begin{equation} \label{eq:vm_model}
C_{jp}(t+\delta t)=\sum_{i\in F_{jp}}\Psi_{i}(t)\omega_{pij} 
\end{equation}

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation} \label{eq:neural_model}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
\noindent where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. The projection strength scaling factor of the afferent project $\gamma_{A}$ was set to 1.5 
based on Stevens \etal\,\citep{Stevens2013} while the values of the lateral excitatory  and inhibitory scaling factors, $\gamma_{E}$
and $\gamma_{I}$ respectively, were varied (see figure \ref{fig:figure1}) to find a balance between excitation and inhibition, and between afferent and lateral influences, to provide robust formation of activity bubbles that allows smooth maps to form.

@@

\noindent where $\xi=\insval{homeostatic_learning_rate}$ is the time constant of the threshold adaptation,
$\mu=\insval{target_activity}$ is a constant defining the target average activity, and $\tilde{\Psi}$ is 
the recent average activity of the unit:

@@

Initial connection field weights are isotropic 2D Gaussians for the lateral excitatory projection and uniformly random within a Gaussian envelope for afferent and lateral inhibitory projections. Specifically, a neuron located at (i, j) will have the following weights in projection $p$:

\begin{equation}
\omega_{ijp} = \frac{1}{Z_{p}} u\, exp(- \frac{x^2+y^2}{2\sigma^2_p} )
\end{equation}

\noindent where (x, y) is the sheet-coordinate location of the presynaptic neuron, $u=1$ for the lateral excitatory projection ($p$=E), and u is a scalar value drawn from a uniform random distribution for the afferent and lateral inhibitory projections ($p$ = A, I),
$\sigma_{p}$ determines the width of the Gaussian in sheet coordinates ($\sigma_{A}=0.27$,$\sigma_{E}=0.035$,$\sigma_{I}=0.035 \sqrt{2}$), 
and $Z_{p}$ is a constant normalizing term that ensures that the total of all weights $\omega_{ijp}$ to neuron $j$ in projection $p$ is 1.0. Weights for each projection are only defined within a specific maximum circular radius $r_p$ ($r_{A}=0.27$,$r_{E}=0.15$,$r_{I}=0.15 \sqrt{2}$).

@@

\noindent where $\beta_{p}$ is the Hebbian learning rate for the connection fields in the two afferent projections from RGC/LGN $p \in \{\mathrm{ON},\mathrm{OFF}\}$. I.e., the afferent weights from RGC/LGN are normalized jointly.  Learning rate parameters are specified as
a fixed value $\iota_{p}=\insval{learning_rate}$ for each projection, and then the unit-specific values used in the equation above are calculated as
$\beta_{p}=\frac{\iota_{p}}{\upsilon_{p}}$, where $\upsilon_{p}$ is the number of connections per connection field in projection $p$.  
The base parameters described here correspond to the first model variant (figure \ref{fig:figure1}). Any modifications of these base paremters in the other two GCAL model variants 2 and 3 (figure \ref{fig:figure2},\ref{fig:figure3}) examined in 
sections \ref{sec:SM2} and \ref{sec:SM3} are then reported in the respective sections.

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelVariants/drawing.png}
\caption{The four considered intra-cortical connectivty variants. As in figure \ref{fig:model_architecture}, in all pannels, the top line represent excitatory and bottom line inhibitory population of neurons. Models (A-C) are using firing rate models of neural units with instantenous 
translation of inputs into membrane-potential (see section \ref{sec:gcal}). 
(A) Variant 1, assuming only local equaly wide excitatory and inhibitory connectivity, and ignoring direct inhibitory to inhibitory interactions. (B) Variant 2, as in A but with added direct inhibitory to inhibitory connections of equal extent as the other excitatory and inhibitory projections. (C) Variant 3, as in A but with added long-range excitatory connections represented based on Buz\'as \etal\,\citep{Buzas2006} as a second wider gaussian. (D) The same as A but modelled with neural units that take into account membrane time-cosntant (see section \ref{sec:rate}).
The transmission delay of the inhibitory to excitatory projections in model 4 was varied in the experiments presneted in section \ref{sec:SM4}
(figure \ref{fig:figure4}).}
\label{fig:model_variants}
\end{figure}

\subsection{Rate model with membrane time constant} \label{sec:rate}

The architecture of this model (model variant 4; figure \ref{fig:model_architecture} and \ref{fig:model_variants}D) is identical to the GCAL models 
(model variants 1 through 3; figure \ref{fig:model_variants}A-C) with the exception of the equations \ref{eq:vm_model} and \ref{eq:neural_model}, which were replaced with equation taking into account the membrane time constant:

\begin{equation}
\tau_{z} \frac{\delta\Psi_{i}}{\delta t} = - \Psi_{i} + \sum_{p}\gamma_{p}\sum_{j\in F_{ip}}f(\Psi_{j}(t-\nu_{p}))\omega_{pji} 
\end{equation}
\noindent where $\tau_{z}$ ($z \in \{E,I\}$) is the membrane time constant of excitatory and inhibitory neurons 
($\tau_{E}$ = \insval{membrane_time_constant_E} ms, $\tau_{I}$ = \insval{membrane_time_constant_I} ms), and $\nu_{p}$ is the transmission delay 
on the projection $p$ ($\nu_{EE}$ = \insval{td_EE}ms, $\nu_{EI}$ = \insval{td_EI} ms,$\nu_{IE}$ was varied (see section \ref{sec:SM4})).
As a consequence the evolution of the network dynamics have to be simulated at higher resolution. We chose the update step to be \insval{rm_update_step}ms, and we let the activity in the model to settle for \insval{rm_settle_time}ms, thus resulting in \insval{rm_settle_steps} settling steps as opposed to the \insval{settle_steps} of the GCAL model. The sub-cortical parametrization of the model is identical to the model variants 1-3 (figure \ref{fig:model_variants}A-C), but because the addition of membrane time constants substantially changes the cortical dynamics
the cortical parameters had to be re-adjusted. Specifically the strength of the thalamo-cortical connections was set to $\gamma_{A}=\insval{w_A}$,
the strength of the lateral excitatory-to-excitatory connections was set to $\gamma_{EE}=\insval{w_EE}$, the strength of the lateral excitatory-to-inhibitory connections was set to $\gamma_{EI}=\insval{w_EI}$, and the strength of the lateral inhibitory-to-excitatory connections was set to $\gamma_{IE}=\insval{w_IE}$.

@@

Model orientation maps are calculated based on the vector average method \citep{CMVC}. We first determine the preferred frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean preferred spatial frequency across all cortical neurons as the
value for the spatial frequency parameter across all subsequent analysis. Next, sine grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180$\degree$), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

In this article we will proceed through multiple gradually more complex models of orientation development, addressing several 
of the major issues with modeling of this phenomena, eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range inhibition can support development of cortical functional organization. We will use two different computational abstractions to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM family of models \citep{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain further in section \ref{sec:SM4}, some simplifications made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.
